\\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates

### 2022-11-25 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

```{r plot-erspc-data, eval = TRUE, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset. A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up. B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. D: The base series, a representative sample of the entire grey area, is represented by the green points."}
# For increased visibility, remove some cases
toRemove <- sample(which(ERSPC$DeadOfPrCa == 1), 
                   round(length(which(ERSPC$DeadOfPrCa == 1))*0.7))
# Create new dataset for illustration
ERSPC_sub <- ERSPC
ERSPC_sub$DeadOfPrCa[toRemove] <- 0

@@

We then compare the risk estimation over the test set. The predicted probabilities for each test set observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier)A. The Kaplan-Meier curve is calculated on the test set only. We see minimal differences between the four approaches across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \@ref(fig:cs3FinalBrier)B, we see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between the three adjusted models are minimal.

@@

```{r cs3FinalBrier, echo = FALSE, eval = TRUE, fig.cap="Comparison of Cox regression (Cox), penalized Cox regression (Pen. Cox), penalized case-base sampling estimation (Pen. CB), and Kaplan-Meier (K-M). (A) Probability of death as a function of follow-up time which is the average of the predicted probabilities for each test set observation. The Kaplan-Meier curve is calculated on the test set only. (B) Brier score as a function of follow-up time, where a lower score corresponds to better performance. We see minimal differences between the four approaches across follow-up-time for the absolute risk curves. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. We see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between the three adjusted models are minimal.", warning = FALSE}
# Combine scores
data_brier <- bind_rows(
  brierCoxKM$Brier$score %>% 
    mutate(model = as.character(model)) %>% 
    filter(model != "Null model"),
  brierPenalized$Brier$score %>% 
    mutate(model = as.character(model)) %>% 
    filter(model != "Null model")
  ) %>% 
  mutate(model = factor(model, levels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))

>>>>

```{r plot-erspc-data, eval = TRUE, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset. A: The gray area can be thought of as N=159,893 infinitely thin horizontal rectangles ordered by length of follow-up. B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. D: The base series, a representative sample of the entire grey area, is represented by the green points."}
# For increased visibility, remove some cases
toRemove <- sample(which(ERSPC$DeadOfPrCa == 1), 
                   round(length(which(ERSPC$DeadOfPrCa == 1))*0.7))
# Create new dataset for illustration
ERSPC_sub <- ERSPC
ERSPC_sub$DeadOfPrCa[toRemove] <- 0

@@

We then compare the risk estimation over the test set. The predicted probabilities for each test set observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier)A. The Kaplan-Meier curve is calculated on the test set only. We see minimal differences between the four approaches across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \@ref(fig:cs3FinalBrier)B, we see that the adjusted models all perform similarly, outperforming the Kaplan-Meier estimate.

@@

```{r cs3FinalBrier, echo = FALSE, eval = TRUE, fig.cap="Comparison of Cox regression (Cox), penalized Cox regression (Pen. Cox), penalized case-base sampling estimation (Pen. CB), and Kaplan-Meier (K-M). (A) Probability of death as a function of follow-up time which is the average of the predicted probabilities for each test set observation. The Kaplan-Meier curve is calculated on the test set only. We see minimal differences between the four approaches across follow-up-time for the absolute risk curves. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. (B) Brier score as a function of follow-up time, where a lower score corresponds to better performance. We see that the adjusted models all perform similarly, outperforming the Kaplan-Meier estimate.", warning = FALSE}
# Combine scores
data_brier <- bind_rows(
  brierCoxKM$Brier$score %>% 
    mutate(model = as.character(model)) %>% 
    filter(model != "Null model"),
  brierPenalized$Brier$score %>% 
    mutate(model = as.character(model)) %>% 
    filter(model != "Null model")
  ) %>% 
  mutate(model = factor(model, levels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))
